---
title: "Create multistate events using the mgus2 data"
author: "Beth Atkinson"
output: html_document
---

```{r setup, include=FALSE}
## load libraries
library(survival)
library(survminer)
library(arsenal)
library(broom)
library(tidyverse)
library(knitr)

## set up options for markdown report
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, collapse=TRUE,
               prompt=TRUE, cache=FALSE, comment=NA, tidy=FALSE)

options(stringsAsFactors=F,
        contrasts=c("contr.treatment","contr.poly"),
        show.signif.stars = FALSE)
```

This exercise is an example of splitting up the observation time into "chunks".

## Description of the veteran cohort.

Randomised trial of two treatment regimens for lung cancer. 

* time:	survival time
* status:	censoring status (0=censor, 1=dead)
* trt:	1=standard 2=test
* celltype:	1=squamous, 2=smallcell, 3=adeno, 4=large
* karno:	Karnofsky performance score (100=good)
* diagtime:	months from diagnosis to randomisation
* age:	in years
* prior:	prior therapy 0=no, 10=yes

```{r}
# look at the dataset
glimpse(veteran)
```

Fit a Cox model - motivation

```{r}
fit1 <- coxph(Surv(time, status) ~ karno, data=veteran)
plot(cox.zph(fit1), lwd=2)
abline(h=0, col=2)
```

## Split data

```{r}
# prior to splitting
veteran$id <- 1:nrow(veteran)
veteran[1:4, c('id','time','status','age','trt')]

vet2 <- survSplit(Surv(time, status) ~., veteran,
                   cut=c(60, 120), episode ="timegp")

# after splitting
vet2 %>% filter(id %in% 1:4) %>% select(id, tstart, time, status,timegp, age, trt)
```

What if you wanted to change age at the beginning of each interval?  How would you do that?

## Run Cox model on new data

```{r}
fit2 <- coxph(Surv(tstart, time, status) ~ 
                   karno:strata(timegp),
              data = vet2)
```

Compare coefficients.

```{r}
# original 
round(coef(fit1), 3)

# within time period
tmp <- round(coef(fit2), 3)
names(tmp) <- c('t0_60','t60_120','t120+')

tmp
```