---
title: "Creating data for multistate analysis using the nafld data"
author: "Beth Atkinson"
output: html_document
---

```{r setup, include=FALSE}
## load libraries
library(survival)
library(survminer)
library(arsenal)
library(broom)
library(tidyverse)
library(knitr)
library(hexbin)

## set up options for markdown report
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, collapse=TRUE,
               prompt=TRUE, cache=FALSE, comment=NA, tidy=FALSE)

options(stringsAsFactors=F,
        contrasts=c("contr.treatment","contr.poly"),
        show.signif.stars = FALSE)
```

This exercise is more complex example of multistate data, created using the `tmerge` function.

## Outline

* A. Allen, Non-alcoholic fatty liver disease incidence and
      impact on metabolic burden and death, a 20 year community study.
      Hepatology 2018, 67:1726--1736.
* The prevalence of non-alcoholic fatty liver disease (NASH) has risen to 24\%.
* Now the most common cause of chronic liver disease.
* Diagnosed with abdominal MRI.
* NASH = NAFLD + inflammation requires biopsy for diagnosis.


```{r}
ncount <- table(sign(nafld1$ntime), useNA="ifany")
n2 <- c(ncount[1]+ ncount[2], ncount[2] + ncount[3])
```

* All NAFLD diagnosis from 1997 to 2014 in Olmsted County, Minnesota.
* Utilize the Rochester Epidemiology Project
* One year delay.
* 4 controls matched on age and sex, then followed forward until the
    analysis date. 
* `r n2[1]` cases of NAFLD and `r n2[2]` controls, with
      `r ncount[2]` subjects that overlap.


## Create Diagram

```{r}
states <- c("No comorbidity", "1 comorbidity", "2 comorbidities", 
            "3 comorbitities", "Death")
cmat <- matrix(0, 5,5)
cmat[,5] <- 1
cmat[1,2] <- cmat[2,3] <- cmat[3,4] <- 1
cmat[1,3] <- cmat[2,4] <- 1.6
cmat[1,4] <- 1.6
dimnames(cmat) <- list(states, states)
statefig(cbind(4,1), cmat)
```

## Data

* `nafld1`: One observation per subject.  Baseline covariates
      plus follow-up time and death.
* `nafld2`: Variables of id, days, test, and value.  Contains
      selected tests and clinical observations.
* `nafld3`: Variables of id, days, and event type.  One observation
      for each outcome: occurrence of NASH, hypertension, diabetes, etc.
* To anonomize patients, all dates have been replaced with ``days
      since first enrollment''.  

Focus is on:

* Metabolic comorbidities are diabetes, hypertension, and dyslipidemia
* Focus on a model with 0, 1, 2, 3, of these + death
* The real work is in building and checking a data set,
      the fits will be easy.
 
```{r}
hexbinplot(I(futime/365.25) ~ age, data=nafld1, colramp=rainbow,
           xlab="Age at entry", ylab="Years of follow-up")
```

## Create analysis data using tmerge

```{r}
keep <- c("id", "age", "male", "bmi", "ntime")
data1 <- tmerge(nafld1[, keep], nafld1, id,  
                death= event(futime, status))
data1 <- tmerge(data1, subset(nafld3, event=="nafld"), id,
                nafld = tdc(days))
data1 <- tmerge(data1, subset(nafld3, event=="diabetes"), id,
                diab= tdc(days), e1= event(days))
data1 <- tmerge(data1, subset(nafld3, event=="htn"), id, 
                htn= tdc(days), e2= event(days))
data1 <- tmerge(data1, subset(nafld3, event=="dyslipidemia"), id, 
                dyslip = tdc(days), e3= event(days))
attr(data1, 'tcount')
```

Check data by looking at someone who has four rows in `data1` (has all 4 events).

```{r}
rowcount <- table(data1$id)
table(rowcount)    # pick someone with 4 rows
subset(data1, id == 135, c(id, age, tstart, tstop, nafld,
                     dyslip, htn, diab, death))
```

Look at the same subject in `nafld3`

```{r}
subset(nafld3, id==135)
```

Create range (death creates the largest range).

```{r}
test <- tmerge(nafld1[, 1:2], nafld1, id,
               death = event(futime, status))
attr(test, "tcount")

subset(test, id==135)
```

Now add in other information

```{r}
test <- tmerge(nafld1[, 1:2], nafld1, id, 
               death = event(futime, status))
test <- tmerge(test, subset(nafld3, event=="nafld"), id,
               nafl = tdc(days))
test <- tmerge(test, subset(nafld3, event=="diabetes"), id,
                diab= tdc(days), e1= event(days))
attr(test, "tcount")
#
subset(test, id==135)
```

Now add in other events

```{r}
test <- tmerge(test, subset(nafld3, event=="htn"), id,
                htn= tdc(days))
attr(test, "tcount")
#
subset(test, id==135)

test <- tmerge(test, subset(nafld3, event=="dyslipidemia"), id,
                lip= tdc(days), e3= event(days))
attr(test, "tcount")
#
subset(test, id==135)
```

Finally, create some additional variables

* `age1`, `age2`: age at start and end of interval
* `cstate`: number of metabolic conditions so far
* `endpoint`: censor, 1mc, 2mc, 3mc, death

```{r}
data1$age1 <- with(data1, age + tstart/365.25)
data1$age2 <- with(data1, age + tstop/365.25)
data1$cstate <- with(data1, diab + htn + dyslip)  # TD cov

tcount <- with(data1, e1 + e2 + e3)
temp2 <- with(data1, ifelse(death, 4, 
              ifelse(tcount ==0, 0, cstate + tcount)))
data1$endpoint <- factor(temp2, 0:4, 
         c("censored", "1mc", "2mc", "3mc", "death"))
data1$cstate <- factor(data1$cstate, 0:3,
                c("0mc", "1mc", "2mc", "3mc"))       
with(data1, table(cstate, endpoint))
```

Check data

```{r}
survcheck(Surv(tstart, tstop, endpoint) ~ male + nafld, data=data1, 
          id=id, istate=cstate)
```


## Fit Models 

```{r}
nfit1 <- coxph(Surv(age1, age2, death) ~ male + nafld,
               data=data1)
nfit2 <- coxph(Surv(age1, age2, death) ~ male + nafld + 
                   as.numeric(cstate),
               data=data1)
nfit3 <- coxph(Surv(age1, age2, death) ~ male + 
                  strata(cstate)/nafld,  data= data1)
nfit4a <- coxph(Surv(age1, age2, endpoint %in% c("1mc", "2mc", "3mc")) ~ 
                      strata(male) + nafld,
               data=data1, subset= (cstate=="0mc"))
nfit4b <- coxph(Surv(age1, age2, endpoint %in% c("2mc", "3mc")) ~ 
                    strata(male) + nafld,
               data=data1, subset= (cstate== "1mc"))
nfit4c <- coxph(Surv(age1, age2, endpoint=="3mc") ~ 
                    strata(male) + nafld,
               data=data1, subset= (cstate=="2mc"))
```

Summarize results from fits

```{r}
mtable <- matrix(NA, 11, 3)
dimnames(mtable) <- list(c("Alive -> Dead", "Alive -> Dead", "","0 MC -> Dead", 
                           "1 MC -> Dead", "2 MC -> Dead", "3 MC -> Dead", "",
                           "0 MC -> 1+ MC", "1 MC -> 2+ MC", "2 MC -> 3 MC"),
                         c("male sex", "NAFLD", "mcount"))
mtable[1, 1:2] <- coef(nfit1)
mtable[2, 1:3] <- coef(nfit2)
mtable[4:7,1:2] <- cbind(coef(nfit3)[1], coef(nfit3)[-1])
mtable[9:11, 1:2] <- rbind(coef(nfit4a), coef(nfit4b), coef(nfit4c))

print(round(exp(mtable), 1), na.print="")
```

* The incremental impact of NAFLD on death decreases
* The impact on conversion to the next comorbid state decreases.
* (Allen adds these to figure 4.)

## Fit Aalen-Johansen curves

```{r }
multi <- survfit(Surv(age1, age2, endpoint) ~ nafld, data=data1,
                istate=cstate, id=id, se=FALSE, start.time=50)
multi$states
plot(multi[,3], col=1:2, xlab='Age', ylab='Probability of 2mc')
legend("topright", legend=names(multi$strata), 
       col=1:2, lty=1, bty='n')
```

