---
title: "Basic Survival Analysis"
author: "Beth Atkinson"
output: html_document
---

```{r setup, include=FALSE}
## load libraries
library(survival)
library(survminer)
library(arsenal)
library(broom)
library(tidyverse)
library(knitr)

## set up options for markdown report
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, collapse=TRUE,
               prompt=TRUE, cache=FALSE, comment=NA, tidy=FALSE)

options(stringsAsFactors=F,
        contrasts=c("contr.treatment","contr.poly"),
        show.signif.stars = FALSE)
```

## Outline

* Create a Kaplan-Meier Curve
  + Use plot function in `survival`
  + Use ggsurvplot in `survminer`
  + Add risktable to plot from `survival` package
  
* Run Cox model

## Explore `ovarian` in the `survival` package.

```{r}
glimpse(ovarian)
```

## Create a Kaplan-Meier object

Use the `survfit` function to create a `survfit` object for further exploration using the endpoint variables `futime` and `fustat`, stratified by **`rx`**.

```{r}
fit <- survfit(Surv(futime, fustat) ~ rx, data=ovarian)
fit
```

* What is the median survival time?
* What is the 25th percentile survival time?

```{r}
quantile(fit, probs=.25)
```

### Plot a KM curve

1. Create plot using the `survival` package
   + Use years as the time scale
   + Limit the x-axis
   + Add labels
   
```{r}
plot(fit, xscale=365, xlim=c(0, 2*365), 
     col=1:2, lty=1:2, 
     xlab='Time since enrollment, years', 
     ylab='Survival probability')
legend('topright', legend=names(fit$strata), 
       col=1:2, lty=1:2, bty='n')
```

2. Repeat plot using the `survminer` package

```{r}
ggsurvplot(fit, xscale=365, xlim=c(0, 2*365), 
           censor=FALSE, break.x.by=365/2,
           risk.table=TRUE,
           xlab='Time since enrollment, years', 
           ylab='Survival probability')
```

3. Add risk table under the plot.survfit R figure

```{r}
# shrink default text size, margins
par(cex=.9, mai=c(.8,1.2,.5,.1))

# select top portion of the plotting window 
par(fig=c(x1=0,x2=1,y1=.2,y2=1))
plot(fit, xscale=365, xlim=c(0, 2*365), 
     col=1:2, lty=1:2, lwd=2,
     xlab='Time since enrollment, years', 
     ylab='Survival probability')
title('Base R, risk table')
legend('topright', legend=names(fit$strata), 
       col=1:2, lty=1:2, bty='n')

# select bottom portion of the plotting window
par(fig=c(0,1,0,.2), new=TRUE, mai=c(.1,1.2,0,.1))
tmp <- summary(fit, times=seq(0,2,.5), scale=365)
plot(c(0,2),c(0,2), axes=FALSE, pch='', xlab='', ylab='', xlim=c(0,2), ylim=c(0,2.5))
text(x=c(tmp$time), y=rep(2:1, each=5), labels=c(tmp$n.risk))
axis(2, at=2:1, c('rx=1','rx=2'), las=1, lty=0)

# set plotting parameters back to default
par(cex=1, mai=c(1.02,.82,.82,.42), fig=c(0,1,0,1)) 
```

### Create a summary by 6 month intervals

```{r}
summary(fit, time=seq(0,2,.5), scale=365)
```

### Create summary using the `arsenal::tableby` function

```{r, results='asis'}
tab <- tableby(rx ~ age + resid.ds + Surv(futime,fustat), data=ovarian)
summary(tab)
```

## Fit a Cox model

Use the covariates age, resid.ds, rx, and ecog.ps.

```{r}
# standard summary from Cox model
cfit <- coxph(Surv(futime,fustat) ~ age + resid.ds + rx + ecog.ps, data=ovarian)
summary(cfit)

# Save results in a dataframe using the tidy function in broom
# Exponentiate and rename the "estimate" to HR (hazard ratio)
tmp <- tidy(cfit, exponentiate=TRUE)
tmp <- tmp %>% rename(HR=estimate) %>%
   select(term, HR, conf.low, conf.high, p.value)

# Transform p-value so that it shows 3 digits or < 0.001
tmp$p.value <- format.pval(round(tmp$p.value, 3), digits=3, eps=.001)

# Pretty table
knitr::kable(tmp, digits=2)
```