---
title: "Create time-dependent covariate using the jasa data"
author: "Beth Atkinson"
output: html_document
---

```{r setup, include=FALSE}
## load libraries
library(survival)
library(survminer)
library(arsenal)
library(broom)
library(tidyverse)
library(knitr)

## set up options for markdown report
knitr::opts_chunk$set(warning = FALSE, message = FALSE, collapse = TRUE)
```

This example highlights how `tmerge` can be used to create multiple events/person.

## Outline

* Chronic Granulotamous Disease (`cgd0`)
  + id, treat, sex, age
  + futime: follow-up time
  + etime1-etime7: up to 7 infection times/subject

* Create a dataset with these variables, 
where `enum` is the interval number/id


|id | tstart | tstop | infect | treat | enum |
|:--|:-------|:------|:-------|:------|:-----|
| 1 | .      | .     | .      | .     | .    |


## Explore `cgd0` dataset 

```{r}
glimpse(cgd0)

# How many people have 7 infections? 
table(!is.na(cgd0$etime7))

cgd0 %>% filter(!is.na(etime7))
```

## Use tmerge to create new dataset

Define follow-up region

```{r}
newcgd <- tmerge(data1=cgd0, data2=cgd0, 
                 id=id, tstop=futime)

attr(newcgd, "tcount")
newcgd[newcgd$id==2, c('id','age','tstart','tstop')]
```

Add in events (`etime1` - `etime7`)

```{r}
newcgd <- tmerge(newcgd, cgd0, id=id,
                 infect=event(etime1), infect=event(etime2),
                 infect=event(etime3), infect=event(etime4),
                 infect=event(etime5), infect=event(etime6),
                 infect=event(etime7))
attr(newcgd, "tcount")
newcgd[newcgd$id==2, c('id','age','tstart','tstop', 'infect')]
```

What is trailing?  Is it ok?

```{r}
cgd0 %>% filter(etime2==futime)

newcgd %>% filter(id==87)
```


Add in `enum` which is a time-dependent covariate.

```{r}
newcgd <- tmerge(newcgd, newcgd, id=id,
                 enum=cumtdc(tstart))

attr(newcgd, "tcount")
newcgd[newcgd$id==2, c('id','age','tstart','tstop', 'infect','enum')]

table(newcgd$enum)
```

## Run a Cox model

```{r, echo=TRUE}
fit <- coxph(Surv(tstart,tstop,infect) ~ treat + steroids + inherit, 
             id=id, data=newcgd)
fit
```

Look at the first infection versus all infections

```{r}
fit0 <- coxph(Surv(tstart,tstop,infect) ~ treat + steroids + inherit, 
              id=id, data=newcgd, subset=enum==1)
round(cbind(first=coef(fit0), all=coef(fit)), 3)
```

How many events were added after the first event?

What happens if you run the `coxph` function without `id=id`?
