---
title: "Create time-dependent covariate using the jasa data"
author: "Beth Atkinson"
output: html_document
---

```{r setup, include=FALSE}
## load libraries
library(survival)
library(survminer)
library(arsenal)
library(broom)
library(tidyverse)
library(knitr)

## set up options for markdown report
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, collapse=TRUE,
               prompt=TRUE, cache=FALSE, comment=NA, tidy=FALSE)

options(stringsAsFactors=F,
        contrasts=c("contr.treatment","contr.poly"),
        show.signif.stars = FALSE)
```

This example highlights how `tmerge` can be used to create multiple events/person and there are intentional gaps where subjects are not at risk for another event. The initial dataset already has multiple obs/subject.

## Outline

1. Create range for when subjects are under observation (`tmerge`)
2. Create event for each infection (`tmerge`)
3. Create intervals where they are not at risk (`tmerge`)
4. Remove intervals where not at risk
5. Add a counter for each person (`tmerge`)
6. Check data and `tcount` attribute


## Use tmerge to create analysis dataset

Quick look at the data.

```{r}
glimpse(rhDNase)

# number obs/id
table(table(rhDNase$id)) 

# number of events (infections)
table(!is.na(rhDNase$ivstart)) 
```

Create range for when subjects are under observation.

```{r}
# create new variable end.tm 
dnase <- rhDNase %>% arrange(id, ivstart) %>%
  mutate(end.tm = as.numeric(end.dt - entry.dt)) 

# 1st obs/id (baseline data)
dnase.b <- dnase %>% distinct(id, .keep_all=TRUE) 
          
dn1 <- tmerge(dnase.b[,c('id','inst','trt','fev')], 
             dnase.b,
             tstop=end.tm, id=id)

dn1[dn1$id==129, ]
```

Create event for each infection.

```{r}
dn2 <- tmerge(dn1, dnase, 
              infect=event(ivstart), id=id)

attr(dn2, "tcount")
dn2[dn2$id==129,]
```

Create intervals where they are not at risk.

```{r}
dn3 <- tmerge(dn2, dnase, 
             no.risk=event(ivstop+6), id=id)

attr(dn3, "tcount")
dn3[dn3$id==129,]

dnase %>% filter(ivstop+6 == end.tm)
dn3 %>% filter(id==413)
```

Remove intervals where not at risk.

```{r}
dn4 <- dn3[dn3$no.risk!=1,]
```

Add a counter for each person.

```{r}
newdnase <- tmerge(dn4, dn4, enum=cumtdc(tstart), id=id)

attr(newdnase, "tcount")
newdnase[newdnase$id==129,]
```

## Run Cox model

```{r}
fit <- coxph(Surv(tstart, tstop, infect) ~ trt + fev, data=newdnase, id=id)
summary(fit)
```