---
title: "Create multistate events using the mgus2 data"
author: "Beth Atkinson"
output: html_document
---

```{r setup, include=FALSE}
## load libraries
library(survival)
library(survminer)
library(arsenal)
library(broom)
library(tidyverse)
library(knitr)

## set up options for markdown report
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, collapse=TRUE,
               prompt=TRUE, cache=FALSE, comment=NA, tidy=FALSE)

options(stringsAsFactors=F,
        contrasts=c("contr.treatment","contr.poly"),
        show.signif.stars = FALSE)
```

This exercise is an example of creating competing risk and multistate data using the `tmerge` function.

## Description of the MGUS cohort.

Monoclonal Gammopathy of Undetermined Significance (MGUS):

* Subjects with a dominant clone in their plasma cell population,
      but without malignancy ($\ge 2$\% of plasma cells in the clone).  
* Normally found incidentally to other tests.
* Should the patient be worried?
* About 1\% per year convert to overt malignancy.
* Essentially independent of age and sex.


* 1384 subjects with monoclonal gammopathy of undetermined significance
      (MGUS)
* R. Kyle, New Engl J Med 346:564-569 (2002)
* Questions
   + Pattern of death and progression
   + Relationship to age, sex, hemoglobin, creatinine,
     and amount of protein in the "spike"

The `mgus2` dataset has two sets of variables that we are interested in. Time and event variables for:

* Progression (i.e., PCM): ptime and pstat.
* Death: futime and death.

```{r}
# look at the dataset
glimpse(mgus2)
```

## Create a Diagram - Competing Risk

```{r}
# Create names for the possible states
states <- c("MGUS", "PCM", "Death")

# Create matrix describing relationship between states
connect <- matrix(0, nrow=3, ncol=3, 
                  dimnames=list(states, states))

# A non-zero element indicates that an arrow should be 
# drawn between state i (row) and state j (column)
connect[1, c(2,3)] <- 1
connect

# Plot
statefig(layout=c(1,2), connect)
```

## Create Data and Analyze for Competing Risk

Only need 1st event for each subject, so we only need 1 obs/person.

```{r}
# time variable will be follow-up time if there is no PCM, 
# and PCM time otherwise
etime <- with(mgus2, ifelse(pstat==0, futime, ptime))

# event variable will be 0 for censor or 2 for death 
# if there is no PCM, and 1 for PCM
event <- with(mgus2, ifelse(pstat==0, 2*death, 1))

# event variable must be a factor for multistate
event <- factor(event, 0:2, 
                labels=c("censor", "pcm", "death"))

# confirm coding makes sense
library(arsenal)
summary(freqlist(~ event+pstat+death, data=mgus2))
```

Plot Aalen-Johansen, probability of event

```{r}
fit <- survfit(Surv(etime, event) ~ sex, data=mgus2)
plot(fit, col=c(1,1,2,2), lty=c(2,1,2,1), xscale=12, 
     xlab="Years post MGUS diagnosis", ylab="Probability")

# short-cut for creating the group labels
temp.label <- c(outer(c("F:","M:"), 
                      c("PCM","death"), paste))
legend("topleft", c(temp.label), 
       col=c(1,1,2,2), lty=c(2,1,2,1), lwd=2, bty='n')

## Double check the legend...
summary(fit, time=20, scale=12)
```

Now subset to PCM event and just plot those curves.

```{r}
fit$states # columns
fit$strata # rows

plot(fit[,2], col=1, lty=2:1,
     xscale=12, lwd=2, 
     xlab="Years post MGUS diagnosis", ylab="Probability, PCM")
legend("topleft", c('F','M'), 
       col=1, lty=2:1, lwd=2, bty='n')
```

Fit a Cox Model

Note that ID is required in the model fit.

```{r}
cfit <- coxph(Surv(etime, event) ~ sex, data=mgus2, id=id)
cfit
```

## Modify the Diagram for Multistate


```{r}
# Create names for the possible states
states <- c("MGUS", "PCM", "Death")
connect <- matrix(0, nrow=3, ncol=3, dimnames=list(states, states))

# A non-zero element indicates that an arrow should be drawn
# between state i (row) and state j (column)
connect[1, c(2,3)] <- 1
connect[2,3] <- 1
connect

# The layout argument is normally a vector of integers indicating how many
# states are shown in each column.  Here, two columns are specified.  In the first
# column there is one state and in the second column there are two states.
statefig(layout=c(1,2), connect)
```

Are there subjects where PCM and death occur at the same time?

```{r}
mgus2 %>% filter(pstat==1, death==1, ptime==futime) %>%
  select(id, ptime, futime, pstat, death)
```

* What to do with 9 subjects who have PCM & death at the same time?
  + Cannot have a time of length 0, so push progression back by 0.1 month.

```{r}
# if subject progresses and death occurs on the same day, 
# subtract .1 month from the progression time
ptemp <- with(mgus2, ifelse(ptime==futime & pstat==1, 
                            ptime-.1, ptime))

# the first call to tmerge sets the time range, 
# so start with the longest times, which are for death
newdata <- tmerge(mgus2, mgus2, id=id, 
                  death=event(futime, death))

# now add additional observations for progressions
newdata <- tmerge(newdata, mgus2, id, 
                  pcm = event(ptemp, pstat))

attr(newdata, "tcount")

# first create a 0,1,2 event variable and make it a factor
temp <- with(newdata, ifelse(death==1, 2, pcm))
newdata$event <- factor(temp, 0:2, 
                        labels=c("censor", "pcm", "death")) 
```

Check the results

```{r}
survcheck(Surv(tstart, tstop, event) ~ sex, 
          data=newdata, id=id)
```

Fit a Cox model

```{r}
cfit <- coxph(Surv(tstart,tstop,event)~sex, data=newdata, id=id)
cfit
```
