---
title: "Create time-dependent covariate using the jasa data"
author: "Beth Atkinson"
output: html_document
---

```{r setup, include=FALSE}
## load libraries
library(survival)
library(survminer)
library(arsenal)
library(broom)
library(tidyverse)
library(knitr)

## set up options for markdown report
opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, collapse=TRUE,
               prompt=TRUE, cache=FALSE, comment=NA, tidy=FALSE)

options(stringsAsFactors=F,
        contrasts=c("contr.treatment","contr.poly"),
        show.signif.stars = FALSE)
```

This exercise is a basic example of creating a time-dependent covariate using the `tmerge` function.

# Outline

* Stanford heart transplant data (`jasa`)
  + wait.time: time before transplant (tx)
  + futime: follow-up time
  + fustat: dead or alive
  + age

* Create dataset with these variables
  
|id | tstart | tstop | death | age |  tx |
|:---|:-----|:------|:-----|:----|:---------|
| 1 | .   | . | . | . |


## Explore `jasa` in the `survival` package.

```{r}
glimpse(jasa)
```

## Use tmerge to create a time-dependent covariate

Create a subject ID variable

```{r}
jasa$id <- 1:nrow(jasa)
```

Use tmerge to define follow-up region

```{r, eval=FALSE}
sdata <- tmerge(jasa, jasa, id=id,
                death = event(futime, fustat))
```

What happened?

```{r}
jasa %>% filter(futime==0) %>% 
  select(id, futime, fustat, wait.time)
```

 => One subject died on the day of entry and (0,0) is an illegal time interval for  `coxph`.  Give the person 1/2 day of follow-up.

```{r}
jasa$futime <- pmax(0.5, jasa$futime)
```

Rerun and look at the tcount attribute

```{r}
sdata <- tmerge(jasa, jasa, id=id,
                death = event(futime, fustat))

attr(sdata, "tcount")
sdata[sdata$id==4, c('id','age','tstart','tstop','death')]
```

Add in wait.time to create an indicator transplant (`tx`) status

```{r}
sdata <- tmerge(sdata, jasa, id=id,
                tx = tdc(wait.time))

attr(sdata, "tcount")
sdata[sdata$id==4, c('id','age','tstart','tstop','death','tx')]
```

What happens if you call the variable `transplant` instead of `tx`?  Why did this happen?

What does "trailing" mean?  Does the data look right to you?  If not, how might you fix it?

```{r}
# Hint: closer look at the data
jasa %>% filter(wait.time==futime) %>% 
         select(id, futime, fustat, wait.time)
```

=> Trailing indicates that the subject died on the same day as their transplant procedure.  The problem was resolved by moving the transplant back 0.5 day.

```{r}
jasa$wait.time <- if_else(jasa$wait.time==jasa$futime,
                          jasa$wait.time - .5,
                          jasa$wait.time)
```

What does leading mean?  Should we be concerned?

Rerun the data creation again.

```{r}
sdata <- tmerge(sdata, jasa, id=id,
                tx=tdc(wait.time))

attr(sdata, 'tcount')
sdata[sdata$id==4, c('id','age','tstart','tstop','death','tx')]
```


## Run a Cox Model

```{r}
fit <- coxph(Surv(tstart, tstop, death) ~ age + tx, data=sdata)
fit
```

