---
title: "Create time-dependent covariates using the pbcseq data"
author: "Beth Atkinson"
output: html_document
---

```{r setup, include=FALSE}
## load libraries
library(survival)
library(survminer)
library(arsenal)
library(broom)
library(tidyverse)
library(knitr)

## set up options for markdown report
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, collapse=TRUE,
               prompt=TRUE, cache=FALSE, comment=NA, tidy=FALSE)

options(stringsAsFactors=F,
        contrasts=c("contr.treatment","contr.poly"),
        show.signif.stars = FALSE)
```

This exercise is a basic example of creating a time-dependent covariates that are changing values using the `tmerge` function.

## Outline

`pbcseq` is from the Mayo Clinic trial in primary biliary cirrhosis (PBC) of the liver conducted between 1974 and 1984. Subjects were randomized to placeobo or D-penicillaimine.

The data `r nrow(pbcseq)` observations with repeated laboratory values + baseline variables.  

## Explore the `pbcseq` dataset from the `survival` package.

The variable `day` indicates the day that the lab variables were measured.

```{r}
glimpse(pbcseq)

## look at 2 people
pbcseq %>% filter(id %in% c(1,5)) %>% 
  select(id,futime,status,trt,day, alk.phos, bili) 

## any missing values?
colSums(is.na(pbcseq))
```

## Create analysis dataset

Create range for when subjects are under observation.

```{r}
# baseline
pbc_b <- pbcseq %>% select(id:sex) %>% distinct()
head(pbc_b)

# Look at status
table(pbc_b$status)
```

In this particular instance, transplant (1) and death (2) can both be treated as death (during this time period, they would have died soon after if they didn't get the transplant).  

```{r}
pbc_b$status2 <- as.numeric(pbc_b$status>0)
```

Set range for each subject. 

```{r}
# set range
newpbc <- tmerge(pbc_b, pbc_b, id=id, 
                 death = event(futime, status2)) 

attr(newpbc, "tcount")
newpbc[newpbc$id==220,]
```

Create new TDC variables

```{r}
newpbc <- tmerge(newpbc, pbcseq, id = id, 
               ascites = tdc(day, ascites),
               hepato = tdc(day, hepato),
               spiders = tdc(day, spiders),
               edema = tdc(day, edema),
               bili = tdc(day, bili), 
               chol = tdc(day, chol),
               albumin = tdc(day, albumin),
               alk.phos = tdc(day, alk.phos),
               ast = tdc(day, ast),
               platelet = tdc(day, platelet),
               protime = tdc(day, protime),
               stage = tdc(day, stage)
               )

attr(newpbc, "tcount")
newpbc[newpbc$id==220,]
```

How do missing values come into play?

* Missing values in time or value from `data2` are ignored with the consequence: "last value carried forward" 

* Default can be changed by adding `options=list(na.rm=FALSE)` to the second call 
    + Any `tdc` calls with a missing time are still ignored, independent
of the na.rm value, since we would not know where to insert them.

What happens if you add this to the `tmerge` call?

## Run Cox model

```{r}
fit <- coxph(Surv(tstart, tstop, death) ~ trt + age + sex + bili, data=newpbc)
summary(fit)

# Re-scale the hazard ratios (age per 10 year increase, alkphos per 1000 increase)
summary(fit, scale=c(1,10,1,1000))
```

# Evaluate Time delay for covariates 

Assume that the impact of changes in the lab values don't happen for 14 days.  Do the results change?

```{r}
newpbc2 <- newpbc <- tmerge(newpbc, pbcseq, id = id, 
               ascites2 = tdc(day, ascites),
               hepato2 = tdc(day, hepato),
               spiders2 = tdc(day, spiders),
               edema2 = tdc(day, edema),
               bili2 = tdc(day, bili), 
               chol2 = tdc(day, chol),
               albumin2 = tdc(day, albumin),
               alk.phos2 = tdc(day, alk.phos),
               ast2 = tdc(day, ast),
               platelet2 = tdc(day, platelet),
               protime2 = tdc(day, protime),
               stage2 = tdc(day, stage),
               options=list(delay=14)
               )
```

Refit the Cox model.

```{r}
fit2 <- coxph(Surv(tstart, tstop, death) ~ trt + age + sex + bili2, data=newpbc2)
summary(fit2)

cbind(orig=coef(fit),
      delay=coef(fit2))
```